---
title: "Testing pleiotropy in multiparental populations"
author: "Frederick Boehm"
date: "April 22, 2019"
bibliography: research.bib
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: '%current%'
      ratio: "4:3"
    css: [default, 'slides.css', xaringan-themer.css]
    chakra: libs/remark-latest.min.js
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(ggalt)
library(knitr)
ratio <- "4:3"
fig_height <- ifelse(ratio=='4:3', 10, 7.5)
fig_width <- 15
curves_height <- ifelse(ratio=='4:3', 580, 540)
curves_width <- ifelse(ratio=='4:3', 800, 1080)
knitr::opts_chunk$set(#dev="svg", 
                      echo=FALSE, 
                      #results="hide",
                      #fig.height=fig_height,
                      #fig.width=fig_width,
                      warning=FALSE, message=FALSE)
xaringan::summon_remark() # download latest version, locally
```


```{r load_refs, echo=FALSE, cache=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'authoryear', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE, 
           longnamesfirst = FALSE,
           max.names = 1,
           bibpunct = c(" (", ") ", ",", ",", " and ", " "))
myBib <- ReadBib("./research.bib", check = FALSE)
```


```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
#mono_light(
#  base_color = "#1c5253",
#  header_font_google = google_font("Josefin Sans"),
#  text_font_google   = google_font("Montserrat", "300", #"300i"),
#  code_font_google   = google_font("Droid Mono")
#)
solarized_light()
```








# Introduction

- 10,000+ traits with RNA sequencing and mass spectrometry 

- Multiparental populations offer high-resolution QTL mapping

- New analysis tools, such as a pleiotropy test for multiparental populations, are needed 


???

- Together, high-dimensional traits and multiparental populations can advance complex trait genetics

---

# `r Citet(myBib, "jiang1995multiple")` test

- Two-parent crosses    

- Applies to two traits that co-map     

- $H_0$: Pleiotropy    

- $H_A$: Two separate QTL    



---

# `r Citet(myBib, "jiang1995multiple")` test


- Perform a two-dimensional two-QTL scan     

    - $vec(Y) = Xvec(B) + vec(E)$     
    
    - Calculate likelihood at each position    

- Calculate likelihood ratio test statistic     

???

- Y: 2 traits  
- X: genotype probs for 2 markers  
- B: founder allele effects  
- E: random errors, iid within each phenotype  
- vec: stack columns

- 2d scan involves all ordered pairs of markers



---

# Multiparental populations

```{r mppfig, echo = FALSE, out.width = "700px", fig.height=fig_height, fig.width = fig_width}
source("R/hs_fig.R")
# https://github.com/kbroman/Talk_JAX2018/blob/master/R/hs_fig.R
```


???


- start with 8 inbred founder lines  
- inbred lines are color-coded  
- each subject represented by one homologous chromosome pair  
- over successive generations of outbreeding, subjects have smaller and smaller contiguous regions from a single inbred line  



---

# Challenges in multiparental populations

- Complex patterns of relatedness    

.soln[Multivariate random effects]

- Multiple founder lines

.soln[Fixed effect for each founder allele]    


???

Two challenges occur at the point of relating phenotypes to genotypes



---

# Test procedure

- Model: $$vec(Y) = Xvec(B) + vec(G) + vec(E)$$ 
    
- Calculate likelihoods at every grid point

???

- $G$: Polygenic random effects that depend on the kinship matrix  
    
- $E$: Random errors
  
  


---

# Test procedure

- Test statistic: $$- \log_{10} \frac{\max (\text{likelihood under pleiotropy})}{\max (\text{likelihood for separate QTL})}$$    

- Parametric bootstrap to get a p-value    


???

We fit the bivariate QTL models at all points on the two-dimensional grid. With the likelihoods of those model fits, we calculate a likelihood ratio test statistic.

Approximate the null distribution with the distribution of bootstrap test statistics

---

# Application

- `r Citet(myBib, c("logan2013high", "recla2014precise"), .opts = list(max.names = 1))` studied 261 Diversity Outbred mice    

- Measured about two dozen behavioral traits

- Two traits map to Chr 8:

    - "hot plate latency" (57 cM)    

    - "percent time in light" (55 cM)


???

We applied our test to freely available data from the Chesler laboratory.


Do they share a single QTL? 

We examined these two traits because they map to the same region

Hot plate latency is a measure of thermal pain sensitivity

Percent time in light is related to measures of anxiety


---

# Allele effects plots  


```{r effects, echo = FALSE, out.width = "500px"}
include_graphics("figs/coefs.svg")
```

???

- X axis is Chr 8 position
- Y axis is allele effect (on the trait)
- 8 lines per figure: one for each founder line


- Distinct allele effects patterns suggest separate QTL



---

# Profile LOD

$$\text{profile LOD}_{\text{trait 1}}(\lambda_1) = max_{\lambda_2}LOD(\lambda_1, \lambda_2)$$

```{r profiles, out.width = "600px", echo = FALSE}
include_graphics("figs/profile.svg", auto_pdf = FALSE)
```

???

The x axis is Chr 8 position

The y axis is LOD

There are three "traces": one for each trait and one for the pleiotropy hypothesis.

The pleiotropy trace derives from the likelihoods for the pleiotropy models (ie, the diagonal on the grid)

The "hot plate latency" profile LOD trace reveals a broad peak from 55 to 65 cM

The "percent time in light" profile LOD trace displays a sharp peak atop a broad region with higher LOD scores. 

The triangles correspond to the univariate QTL peak positions, while the circles are the profile LOD peak positions.

The two profile LOD traces achieve the same maximum value, which is the test statistic value. 



---

# Test results


- $\log_{10} \Lambda = 1.2$

- $p = 0.11$ (1000 bootstrap samples)

???

- Weak evidence for two separate QTL

    - One QTL affects "distance traveled in light"    
    
    - Second QTL affects "hot plate latency"


---

# qtl2pleio R package

- Functions for $d$-variate, $d$-QTL scan & profile LOD plots

- Uses C++ for matrix calculations (via Rcpp and RcppEigen)

- Uses `gemma2` R implementation of GEMMA EM algorithm for multivariate random effects

- Unit tests, vignettes, and version control


???

- unit tests to ensure accurate calculations

- vignettes on analyzing a pair of traits & on cluster computing for bootstrap test





---

# References



```{r bib, results = 'asis'}
foo <- PrintBibliography(myBib, .opts = list(style = "markdown"), start = 1, end = 3)
```




---

# References

```{r bib2, results = 'asis'}
PrintBibliography(myBib, .opts = list(style = "markdown"), start = 4, end = 6)
```

